Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization

### 2014-08-07 TOMS_Paper/assembly.tex
Rewrite BLAS and add poisson

\subsection{Exposing Linear Alegbra Operations}
In~\cite{Luporini}, we introduced the idea of transforming the element matrix evaluation into a sequence of calls to highly-optimized dense matrices multiply routines (henceforth DGEMM), for instance MKL or ATLAS BLAS. We compared COFFEE's optimizations with hand-written MKL-based kernels, showing how the small sizes of the involved arrays impair DGEMM calls. The study was conducted only in the case of a specific Helmholtz equation. There are cases, however, in which tabulated basis functions sizes grow up to a point for which certain BLAS implementations allow to achieve a better performance. For example, this may be the case of relatively-high polynomial orders or when the form is characterized by the presence of a number of pre-multiplying functions. To explore these extreme yet meaningful scenarios, we have developed an algorithm to reduce any generic element matrix evaluation to a sequence of DGEMM calls.

The algorithm is shown in Figure~\ref{fig:algo-dgemm}. By using the rewrite rules in Figure~\ref{fig:rewrite-rules}, it can be proved that the Expression Rewriter can always reduce an assembly expression to a summation, over each quadrature point, of outer products along the test and trial functions. Each outer product is then isolated, i.e. the assembly expression is split into chunks, each chunk representing an outer product. Terms in the bodies of the various enclosing loops (e.g. coefficients evaluation at quadrature point, temporaries introduced by rewriting the assembly expression) are vector-expanded and hoisted outside of the loop nest, as explained in Section~\ref{sec:vect-exp}. This implies that the loop nest is now perfect; that is, there is no intervening code among the various loops. The element matrix evaluation has now become a sequence of perfect loop nests, each loop nest computing a dense matrix-matrix multiply between temporaries hoisted by the Expression Rewriter or tabulated basis functions. Eventually, the storage layout of the involved operands is changed so as to be conforming to the BLAS interface (e.g. two dimensional arrays are flatten as one dimensional arrays). The translation of each loop nest into a DGEMM call is the last, straightforward step.

>>>>

\subsection{Exposing Linear Alegbra Operations}
In~\cite{Luporini}, we introduced the idea of transforming the element matrix evaluation into a sequence of calls to highly-optimized dense matrix-matrix multiply routines (henceforth DGEMM), for instance MKL or ATLAS BLAS. We compared COFFEE's optimizations with hand-written MKL-based kernels, showing how the small sizes of the involved arrays impair DGEMM routines, which are usually tuned for large arrays. The study was conducted only in the case of a specific Helmholtz problem. There are scenarios, however, in which tabulated basis functions sizes grow up to a point for which turning to BLAS may actually lead to better performance. For example, this can happen when relatively-high polynomial orders are used or if coefficients are present in the form, since they are both responsible for the size of tabulated basis functions. To explore these scenarios, we have developed an algorithm that reduces any assembly expression evaluation to a sequence of DGEMM calls.

Here, we informally provide the main steps of the algorithm. By fully applying the rewrite rules in Figure~\ref{fig:rewrite-rules}, an assembly expression is reduced to a summation, over each quadrature point, of outer products along the test and trial functions. Each outer product is then isolated, i.e. the assembly expression is split into chunks, each chunk representing an outer product over test and trial functions. Statements in the bodies of the surrounding loops (e.g. coefficients evaluation at a quadrature point, temporaries introduced by expression rewriting) are vector-expanded and hoisted completely outside of the loop nest, similarly to what we have described in Section~\ref{sec:rewriting-rules}. This renders perfect the loop nest; that is, there is no intervening code among the various loops. The element matrix evaluation has now become a sequence of dense matrix-matrix multiplies (transposition aside)
\begin{equation*}
A_{jk} = \sum_{i} x_{0_{ij}} \cdot y_{0_{ik}} + \sum_{i} x_{1_{ij}} \cdot y_{1_{ik}} + ...
\end{equation*}
where $x0, x1, y0, y1, ...$ are tabulated basis functions or vector-expanded temporaries introduced at expression rewriting time. Eventually, the storage layout of the involved operands is changed so as to be conforming to the BLAS interface (e.g. two dimensional arrays are flatten as one dimensional arrays). The translation into a sequence of DGEMM calls is the last, straightforward step.
